---
title: "purple air data exploration"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```

## Load in CSV file 

- Data is read in with `read.csv()`
- Summarize all columns with `summary()`
- See how many different sensors are and what they are labelled with `unique()`

```{r}
# load in and check out data 
purp_air <- read.csv("all_data_12324.csv")

# get summary info 
summary(purp_air)

# check how many different sensors there are and their name 
unique(purp_air$sensor)

# check time step
head(purp_air$time_stamp)


```

4 different sensors are included with prefix "CA-"

only column with NA's is `voc`, with 223 NAs 

## Tidy data frame 

- remove 'X' column 
- add column for `date` which turns the `time_stamp` column into date format 
- aggregate to daily values because every 2 hours is a lot of data! 

```{r}
# edit date to timestamp 
purp_tidy <- purp_air %>% 
  select(-X) %>% 
  mutate(date = date(time_stamp)) %>%
  select(-time_stamp) %>% 
  group_by(date, sensor) %>% 
  summarize_all(mean, na.rm=T)

# check new 
head(purp_tidy)

nrow(purp_tidy)
```

## Define columns 

- `time_stamp`
- `humidity`
- `voc`
- `pm1.0_atm`
- `pm2.5_atm`
- `pm10.0_atm`
- `sensor`

## Preliminary plots 

Creating time series with `ggplot`


```{r}
ggplot(purp_tidy, aes(x=date, y=pm10.0_atm, col=sensor)) +
  geom_point() + ggtitle("PM10")

ggplot(purp_tidy, aes(x=date, y=pm2.5_atm, col=sensor)) +
  geom_point() +
  facet_grid('sensor') +
  geom_hline(aes(yintercept=35)) + ggtitle("PM2.5")

ggplot(purp_tidy, aes(x=date, y=pm10.0_atm, col=sensor)) +
  geom_point() +
  facet_grid('sensor') + ggtitle("PM10")


ggplot(purp_tidy, aes(x=date, y=voc, col=sensor)) +
  geom_point() +
  facet_grid('sensor') + ggtitle("VOC")

ggplot(purp_tidy, aes(x=sensor, y=pm2.5_atm, col=sensor)) +
  geom_boxplot() + ggtitle("PM2.5")

```

assuming units are in $\mu g/m^3$

## what are levels of 'AQI'

- create new column above or below threshold 
- check AQI standards 
- plot with lines to mark thresholds
- discrete v. continuous data 
